% !TEX program = xelatex
% Character n-gram Embeddings for Classical Syriac
% Build:  xelatex main && bibtex main && xelatex main && xelatex main
% Requires XeLaTeX + a Syriac-capable font (Noto Sans Syriac). If that font is
% absent, comment out the \newfontfamily\syriacfont line and the \syr macro will
% fall back to the main font (Syriac may not render, transliteration still does).

\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{microtype}
\usepackage[round]{natbib}
\usepackage[hidelinks]{hyperref}
\usepackage{xcolor}

% --- Fonts / Syriac support (XeLaTeX) ------------------------------------ %
\usepackage{fontspec}
% Right-to-left Syriac font. If "Noto Sans Syriac" is not installed, comment the
% next line out; the macro below then falls back to the main font, and the
% accompanying transliteration always renders regardless.
\newfontfamily\syriacfont[Script=Syriac]{Noto Sans Syriac}
\providecommand{\syriacfont}{}            % no-op fallback if the line above is removed
\providecommand{\RL}[1]{#1}               % overwritten by the bidi package below

% \syr{script}{transliteration}{gloss}: native Syriac + translit + English,
% so readers who cannot read the script always have an anchor.
\newcommand{\syrscript}[1]{{\syriacfont\RL{#1}}}
\newcommand{\syr}[3]{\syrscript{#1}~(\textit{#2}, `#3')}
\newcommand{\syrt}[2]{\syrscript{#1}~(\textit{#2})}

\newcommand{\auc}{\textsc{auc}}
\newcommand{\todo}[1]{\textcolor{red}{[#1]}}

% The bidi package must be loaded LAST (after hyperref and other packages).
\usepackage{bidi}

\newcommand{\auc}{\textsc{auc}}
\newcommand{\todo}[1]{\textcolor{red}{[#1]}}

\title{Character $n$-gram Embeddings for Classical Syriac:\\
Subword Modeling of a Templatic Low-Resource Language,\\
with an Application to Authorship Analysis}

\author{Anonymous\\ \texttt{TODO: affiliation} \and Anonymous\\ \texttt{TODO: affiliation}}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
Classical Syriac is a major literary language of late antiquity that remains
severely under-resourced in NLP, yet it raises rich, unresolved questions of
authorship. We train a character $n$-gram \textsc{FastText} model on the Digital
Syriac Corpus (632 texts; 2.18M tokens) and argue that subword representations
are the appropriate inductive bias for a templatic, root-and-pattern language
whose vocabulary is 50\% \emph{hapax legomena}. Intrinsically, the embeddings are
morphologically coherent---forms sharing a triconsonantal root are closer than
semantically adjacent forms built on different roots---and a subword ablation
against \textsc{word2vec} shows the gap is largest for rare and out-of-vocabulary
forms. We then apply the embeddings to authorship: averaged document vectors
separate same- from cross-author text pairs with \auc{} up to $0.90$
(author-cluster bootstrap 95\% CIs; stable across five seeds, $0.886\pm0.007$),
validated by a negative control in which a single author split in two yields
chance \auc{} ($\approx 0.51$). We find that mean-centering to remove a dominant
common component is essential (it lifts \auc{} from $0.73$ to $0.89$), compare
against Burrows's Delta and against from-scratch byte- and character-level neural
language models, quantify a length-dependent genre confound and show the signal
survives genre-matched testing ($0.900\to0.883$), and apply the pipeline to three
disputed or pseudonymous works with historically sensible results. A small
supervised verification head trained on the embeddings and evaluated
leave-one-author-out gives the best separation (\auc{} $0.97$). Notably, the
subword advantage is largest \emph{intrinsically}---on rare and out-of-vocabulary
forms---while at the document level a plain \textsc{word2vec} is competitive.
Code, seeds, and trained models are released.
\end{abstract}

% ===================================================================== %
\section{Introduction}
\label{sec:intro}

Classical Syriac---a dialect of Aramaic and a principal language of eastern
Christianity---preserves a vast literature, much of it anonymous, pseudonymous,
or of contested authorship. Computational stylometry could help, but Syriac is
low-resource: there is no large pretrained language model, limited annotated
data, and the writing system carries optional diacritics that fragment surface
forms.

Syriac morphology is \emph{templatic} (root-and-pattern): most words are built by
interleaving a consonantal root, typically three consonants, with vocalic and
affixal patterns. As a result, forms that share a root---e.g.\
\syr{\u0721\u0720\u071f\u0710}{malk\=a}{king} and
\syr{\u0721\u0720\u071f\u0718\u072c\u0710}{malk\=ut\=a}{kingdom}---are
semantically and morphologically related but orthographically distinct. This,
together with heavy inflection and a long tail of rare forms, motivates a
\emph{subword} model: character $n$-grams let related forms share parameters and
let unseen forms receive a vector at all.

This paper makes a dual contribution---a representation-learning study and a
digital-humanities application.

\paragraph{Contributions.}
\begin{enumerate}
  \item A released character $n$-gram \textsc{FastText} model for Classical
  Syriac (2.18M tokens), with \texttt{min\_count}=1 so that every form, including
  the 50\% of the vocabulary that is hapax, receives a representation
  (\S\ref{sec:model}).
  \item Intrinsic validation: a morphological-coherence probe and a subword
  ablation against \textsc{word2vec}, showing the subword advantage grows for
  rare forms, plus an out-of-vocabulary generalization test
  (\S\ref{sec:results-model}).
  \item A representation bake-off on identical data and evaluation---char
  $n$-gram \textsc{FastText} vs.\ \textsc{word2vec} vs.\ Burrows's Delta vs.\ a
  from-scratch byte-level recurrent LM vs.\ a tiny character Transformer vs.\ a
  small supervised authorship-verification head, the last evaluated
  leave-one-author-out (\S\ref{sec:results-style}).
  \item An embedding-geometry finding: averaged document vectors are strongly
  anisotropic, and label-free common-component removal is necessary for the
  authorship signal to emerge---it lifts \auc{} from $0.73$ to $0.89$
  (\S\ref{sec:model}, \S\ref{sec:results-style}).
  \item A robust same-/cross-author signal (centered \auc{} up to $0.90$,
  bootstrap CIs, multi-seed) validated by a negative control
  (\S\ref{sec:results-style}).
  \item A quantified, length-dependent genre confound, and evidence that the
  authorship signal survives genre-matched testing
  (\S\ref{sec:results-style}).
  \item Attribution of three disputed/pseudonymous works with historically
  sensible outcomes (\S\ref{sec:results-style}).
\end{enumerate}

% ===================================================================== %
\section{Related Work}
\label{sec:related}

\paragraph{Subword and character embeddings.}
\textsc{FastText} represents a word as a bag of character $n$-grams
\cite{bojanowski2017enriching}, extending \textsc{word2vec}
\cite{mikolov2013distributed} to compose vectors for unseen forms---attractive
for morphologically rich and low-resource languages. Character-aware neural
models \cite{kim2016character} and token-free byte models such as ByT5
\cite{xue2022byt5} pursue the same goal with neural architectures
\cite{vaswani2017attention}.

\paragraph{Embedding geometry.}
Averaged embeddings are anisotropic: they share a dominant common direction that
inflates cosine similarities. ``All-but-the-top'' postprocessing
\cite{mu2018allbutthetop} and analyses of contextual geometry
\cite{ethayarajh2019contextual} motivate the mean-centering we apply.

\paragraph{Stylometry and authorship.}
Burrows's Delta \cite{burrows2002delta}, with its geometric and probabilistic
interpretations \cite{argamon2008interpreting,evert2017understanding}, is the
standard baseline; function words are classic markers
\cite{kestemont2014function}, and the field is surveyed in
\cite{stamatatos2009survey}. Topic can confound authorship
\cite{seroussi2014authorship}; we test an analogous genre confound. We assess
significance with the cluster bootstrap \cite{efron1986bootstrap}.

\paragraph{Syriac and digital classics.}
For the language and its study see \cite{brock2006introduction,butts2019syriac};
our data is the Digital Syriac Corpus \cite{digitalsyriaccorpus}. Tooling:
\textsc{gensim} \cite{rehurek2010gensim} and \textsc{PyTorch}
\cite{paszke2019pytorch}.

% ===================================================================== %
\section{Corpus and Preprocessing}
\label{sec:corpus}

We use the Digital Syriac Corpus (the \texttt{srophe/syriac-corpus} TEI/XML
release): 632 texts, one author credited per text for 631 of them.
Table~\ref{tab:corpus} summarizes the corpus. The most striking property for a
representation model is the vocabulary's long tail: half of all distinct forms
occur exactly once.

\input{tables/T1_corpus}

\paragraph{Tokenization.}
We extract maximal runs of Syriac letters (U+0710--U+074F) and combining marks,
requiring at least one letter, and discarding invisible joiners and bidi
controls. By default we \emph{normalize} by stripping combining diacritics
(seyame, vowel points), which aligns inflected forms by their consonantal
skeleton; we report this setting throughout.

\paragraph{Metadata caveats.}
Genre is not encoded in these headers; we recover an approximate genre label from
the series title (\S\ref{sec:results-style}). Texts marked \emph{Anonymous}
collapse to a single non-author and are excluded from author-level analyses.
Authors named without a \texttt{syriaca.org} identifier are merged into their
URI-identified counterpart by name.

% ===================================================================== %
\section{Model and Representations}
\label{sec:model}

\paragraph{FastText.}
We train a skip-gram \textsc{FastText} model: 100 dimensions, character $n$-grams
of length 2--5, $2\times10^{5}$ hash buckets, window 5, 10 epochs, and crucially
\texttt{min\_count}=1 so every form is in the vocabulary. Each TEI text is one
``sentence.'' The rationale is templatic morphology (\S\ref{sec:intro}): shared
character $n$-grams tie together forms of a common root, and the same mechanism
synthesizes vectors for forms unseen in training.

\paragraph{Document and word vectors.}
A document is the frequency-weighted mean of its word vectors. For stylometry we
also consider a \emph{function-word} variant restricted to the 200 most frequent
forms, the topic-independent markers of classical stylometry.

\paragraph{Anisotropy and mean-centering.}
Averaged document vectors share a large common component: raw cosine
similarities cluster near 1 even across authors. We therefore subtract the
corpus-mean document vector (a label-free, unsupervised step) before computing
cosines. \S\ref{sec:results-style} shows this is essential rather than cosmetic.

\paragraph{Neural baselines.}
As architectural contrasts we train two tiny causal language models from scratch
on the same corpus: a byte-level LSTM (\textsc{byte-LM}) over UTF-8 bytes, and a
small character Transformer (\textsc{char-Transformer}) over Unicode codepoints.
Both yield a document vector by mean-pooling final-layer hidden states, and a
word vector by the same pooling over an isolated form, so they enter the same
intrinsic and downstream evaluations. On 2.18M tokens these models are
deliberately small and data-limited; they probe architecture, not scale.

% ===================================================================== %
\section{Stylometry Methods}
\label{sec:methods-style}

\paragraph{Separation \auc{}.}
Given document vectors and author labels, we score every text pair by cosine
similarity and report
$\auc{} = \Pr(\text{same-author pair more similar than cross-author pair})$,
computed via the Mann--Whitney statistic; $0.5$ is chance. We also report
Cohen's $d$ between the two pair-similarity distributions.

\paragraph{Burrows's Delta.}
For each text we form relative frequencies of the top-$k$ most frequent corpus
forms, $z$-score each feature across the corpus, and take the mean absolute
$z$-difference (Manhattan distance) as the Delta distance.

\paragraph{Attribution.}
We attribute a text to the nearest author \emph{centroid}---cosine for the
embeddings, mean Delta for Delta---under strict leave-one-out: the held-out text
is removed from its own author's centroid before scoring. We report top-1 and
top-3 accuracy. As a learned alternative we also train a small
supervised-contrastive projection \citep{khosla2020supcon} (a one-hidden-layer
MLP) on the document vectors so that same-author texts are pulled together; it is
evaluated \emph{leave-one-author-out} (the head is trained only on other
authors), so its embeddings are out-of-sample and comparable to the unsupervised
methods.

\paragraph{Uncertainty.}
Because text pairs are dependent, we bootstrap over \emph{authors} (clusters):
resampling author groups with replacement ($B=1000$) and recomputing \auc{}
gives percentile 95\% confidence intervals. We additionally retrain
\textsc{FastText} under five seeds to report variance.

% ===================================================================== %
\section{Results: The Model}
\label{sec:results-model}

\paragraph{Morphological coherence and the subword ablation.}
Table~\ref{tab:morph} reports, for a fixed set of 8 corpus-verified root families
(49 related pairs), the mean cosine between root-sharing forms and a
frequency-matched control on a different root. FastText shows the largest margin
($+0.32$ frequent, $+0.29$ rare). The decisive contrast is with the no-subword
\textsc{word2vec} baseline, whose margin is smaller and \emph{degrades for rare
forms} ($+0.125\to+0.076$; accuracy $0.85\to0.77$), while FastText stays robust
($0.96\to0.95$). The neural models rank related above control but with cosines
compressed by anisotropy.

\input{tables/T3a_morph}

\paragraph{Out-of-vocabulary generalization.}
With models trained on a 90\% file split, 9{,}023 forms (23.3\% of the held-out
vocabulary) are truly unseen. FastText synthesizes a vector for \emph{every} one
from its character $n$-grams (\textsc{word2vec}: none), and for a 300-form sample
the nearest in-vocabulary neighbour shares the three-consonant root prefix 56\% of
the time (Table~\ref{tab:oov})---the synthesized vectors are morphologically
coherent.

\input{tables/T3b_oov}

\paragraph{Hyperparameter sensitivity.}
The morphology margin and downstream \auc{} are stable across $n$-gram range,
dimension, and skip-gram vs.\ CBOW (Table~\ref{tab:hyper}; \auc{} in
$[0.888,0.916]$), so the results are not an artefact of a particular setting.

\input{tables/T4_hyper}

\paragraph{Neural LM quality.}
Table~\ref{tab:lm} confirms the from-scratch LMs learn Syriac: held-out
bits-per-byte of $0.86$ (byte-LM) and $0.82$ (char-Transformer), far below the
8-bpb uniform-byte baseline.

\input{tables/TLM_lm}

% ===================================================================== %
\section{Results: Authorship}
\label{sec:results-style}

\paragraph{Same- vs.\ cross-author separation.}
Table~\ref{tab:sep} reports \auc{} with author-cluster bootstrap CIs for the full
and restricted cohorts, all-words and function-words. Mean-centered \auc{}
reaches $0.885$ (full) and $0.900$ (restricted, all words); the raw (uncentered)
figures are far lower (e.g.\ $0.73$ vs.\ $0.89$ full), demonstrating that the
anisotropic common component must be removed. Across five training seeds the
restricted \auc{} is $0.886\pm0.007$, so the signal is stable.

\input{tables/T5_sep}

\paragraph{Negative control.}
Splitting a single author (Ephrem) into two pseudo-authors over 20 random halves
yields mean \auc{} $\approx 0.51$: the pipeline finds no boundary where there is
none, so the cross-author signal is genuine.

\paragraph{Representation bake-off.}
Table~\ref{tab:bakeoff} compares all representations on \auc{} and attribution.
Among \emph{unsupervised} methods the count-based ones are strongest and
statistically comparable---\textsc{word2vec} ($0.946$), FastText ($0.915$), and
Delta ($0.907$) at the 1000-token floor---with overlapping CIs; document-level
mean pooling washes out FastText's subword advantage, which is intrinsic rather
than downstream. The tiny from-scratch neural LMs are data-limited, the
char-Transformer ($0.845$) ahead of the byte-LM ($0.762$). A small
\emph{supervised} contrastive AV head trained on the FastText vectors and
evaluated leave-one-author-out gives the best separation \auc{} ($0.966$),
confirming that a learned metric on top of the embeddings helps; a deliberately
leaky variant that trains on the test author reaches $0.999$, so the LOAO number
is a genuine out-of-sample gain. The head does not improve closed-set
attribution (top-1 $0.922$), as its objective targets pairwise verification, not
nearest-centroid classification.

\input{tables/T6_bakeoff}

\paragraph{Genre confound.}
Within Ephrem, hymns vs.\ prose separate weakly when short hymns dominate but
strongly at matched length (\auc{} $0.59\to0.86$); genre/register is thus a
real confound. Yet restricting cross-author comparisons to same-genre pairs
barely lowers separation (Table~\ref{tab:genre}; $0.900\to0.883$), so the
authorship signal is not merely a genre effect.

\input{tables/T7_genre}

\paragraph{Disputed texts.}
Table~\ref{tab:disputed} applies the attributor (leave-one-out top-1 $0.93$ on
known authors) to three held-out cases. The letter transmitted under Ephrem's
name ranks him \emph{last} of eleven and at the 0th percentile of his genuine
texts; the Pseudo-Clementines cluster with translated works (Paul, Eusebius)
rather than with each other; and the Chronicle of Zuqnin is nearest Eusebius, a
historiographic register.

\input{tables/T8_disputed}

% ===================================================================== %
\section{Discussion}
\label{sec:discussion}
The results separate two questions that are often conflated. \emph{Where does the
subword inductive bias help?} Intrinsically and on the long tail: FastText's
morphological margin is largest for rare forms, and it alone assigns coherent
vectors to the 23\% of held-out forms that are out-of-vocabulary. \emph{What
wins the document-level stylometry task?} Here a plain \textsc{word2vec} is
competitive with, even slightly ahead of, FastText: averaging hundreds of word
vectors per document washes out the subword detail, and frequency-profile methods
(Delta) remain strong. We report this honestly rather than overclaiming for
subword embeddings. The authorship signal itself is genuine---it passes a
negative control and survives genre matching---and reflects style rather than
topic alone. The Pseudo-Clementine result is a reminder that translated texts
carry the translator's, not the named author's, fingerprint. Finally, the tiny
from-scratch neural LMs trail the count-based methods: on 2.18M tokens they are
data-starved, which is itself the low-resource condition this language presents.

% ===================================================================== %
\section{Limitations}
\label{sec:limitations}
A single corpus and a small author pool (11--13 in the restricted cohort); the
disputed-text studies are illustrative, without expert-adjudicated gold labels;
the neural baselines are tiny and from-scratch (no large pretrained Syriac LM
exists), so they probe architecture under data scarcity, not an upper bound; the
supervised AV head is likewise trained on only ten-odd authors, so its margin
should be read as a lower bound that more authors would likely widen; and the
genre classifier is an approximate series-title heuristic.

% ===================================================================== %
\section{Conclusion}
\label{sec:conclusion}
We presented a character $n$-gram \textsc{FastText} model for Classical Syriac
and a stylometric study built on it. The subword bias is well suited to a
templatic, hapax-heavy, low-resource language---most clearly on rare and unseen
forms---and the resulting document vectors carry a robust, controlled authorship
signal that is competitive with Burrows's Delta and that yields historically
sensible readings of disputed texts. Future work includes per-stanza
representations for short verse, supervised authorship verification, larger
pretrained byte-level models, and further disputed dossiers.

\paragraph{Reproducibility.}
All experiments are seeded and driven by released scripts
(\texttt{fasttext\_model.py}, \texttt{stylometry.py}, \texttt{authorship.py},
\texttt{nn\_baselines.py}, \texttt{paper\_experiments.py}); the corpus is
CC~BY~4.0 \cite{digitalsyriaccorpus}; trained models are released.

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
